Seminar: Programming models and Code Generation\\ Winter Term 2014/2015\\ {\bf Quasi deterministic parallel programming using LVars}

### 2015-01-09 draft.tex
new

\section{LVars and LVish}
\subsection{LVars}
An LVar is a memory location that can be shared between multiple threads whose contents can only grow bigger over time, for some definition of “bigger” that the user of the LVar gets to specify. \par
LVars, short for lattice variables, are a generalization of IVars, which are variables that can only be written to once. (The “I” is for “immutable”.) LVars, on the other hand, allow multiple writes, as long as those writes are monotonically increasing with respect to a user-specified lattice. Meanwhile, LVar reads have to specify a threshold set of possible states. A read will block until one of the states in the threshold set has been reached, and then return that state. Information can never be removed and the order in which it is added is not observable. IVars turn out to be a special case of LVars. 
The property of monotonically increasing write operations is enforced by the language. The semantics of $put$ says that the write to a location is the least upper bound of the new value and the current value. 
\begin{verbatim}
let par _ = put num 3
        _ = put num 2
        x = get num
  in x  /*could be 2 or 3*/
\end{verbatim}
Monotonic writes alone aren't enough to ensure determinism. The above program is non-deterministic where num's notion of “bigger” is $\geq$. Despite writes always being monotonically increasing. $Get$ might read either $2$ or $3$, depending on the order of $put$ or $get$ operations determined by the scheduler. With an LVar model this behavior changes.\par
\begin{itemize}
\item To maintain determinism, the LVars model restricts the $get$ operation. Instead of reading the LVar’s exact value, it reads one of a set\{threshold set\} of lower bounds on its value that are given as an extra argument to $get$ not shown in the example. The $get$ operation will block until the LVar reaches a value that is at or above one of those lower bounds. For the above example, the threshold set will just be \{$3$\}. $get$ operation is called a threshold read, because the value it returns is the threshold we have crossed.
\item Once num value is atleast $3$, it will remain at or above $3$ forever. As long as we only access them through the $put$ or $get$ interface, LVars are thread-safe: we can safely share them between threads without introducing unintentional non-determinism. That is, $puts$ and $gets$ can happen in any order, without changing the value that a program evaluates to.
\end{itemize}
%\begin{center}
%\textit{Monotonically increasing writes + threshold reads = deterministic parallelism}
%\end{center}
The resulting model is general enough to subsume the existing single-assignment model. The user-specified set just needs to be a \textit{join-semilattice}, which is a partially ordered set in which every two elements have a least upper bound.

@@

Par type constructor($\ast$ is the normal haskell type):

@@

Internally, the Par monad represents computations in continuation-passing style, in terms of their interpretation in the IO monad.

@@

In an attempt towards creating better deterministic parallel programming models, Lvish has extended the existing theme of monotonicity to provide better applicability, usability, performance and a single generic model that subsumes many of the single assignment models, thereby representing a larger class of problems which can be efficiently parallelized. The performance with regards to certain benchmarks are an evidence of the same. At the end of the previous section, some of the drawbacks and a proposed enhancement were highlighted from a recent paper.

>>>>

\section{LVars and LVish}
\subsection{LVars}
An LVar is a memory location that can be shared between multiple threads whose contents can only grow bigger over time, for some definition of “bigger” that the user of the LVar gets to specify. \par
LVars, short for lattice variables, are a generalization of IVars, which are variables that can only be written to once. (The “I” is for “immutable”.) LVars, on the other hand, allow multiple writes, as long as those writes are monotonically increasing with respect to a user-specified lattice. Meanwhile, LVar reads have to specify a threshold set of possible states. A read will block until one of the states in the threshold set has been reached, and then return that state. Information can never be removed and the order in which it is added is not observable. IVars turn out to be a special case of LVars. 
The property of monotonically increasing write operations is enforced by the language. The semantics of $put$ says that the write to a location is the least upper bound of the new value and the current value. 
\begin{verbatim}
let par _ = put num 3
        _ = put num 2
        x = get num
  in x  /could be 2 or 3/
\end{verbatim}
Monotonic writes alone aren't enough to ensure determinism. The above program is non-deterministic where num's notion of “bigger” is $\geq$. Despite writes always being monotonically increasing. $Get$ might read either $2$ or $3$, depending on the order of $put$ or $get$ operations determined by the scheduler. With an LVar model this behavior changes.\par
\begin{itemize}
\item To maintain determinism, the LVars model restricts the $get$ operation. Instead of reading the LVar’s exact value, it reads one of a set\{threshold set\} of lower bounds on its value that are given as an extra argument to $get$ not shown in the example. The $get$ operation will block until the LVar reaches a value that is at or above one of those lower bounds. For the above example, the threshold set will just be \{$3$\}. $get$ operation is called a threshold read, because the value it returns is the threshold we have crossed.
\item Once num value is atleast $3$, it will remain at or above $3$ forever. As long as we only access them through the $put$ or $get$ interface, LVars are thread-safe: we can safely share them between threads without introducing unintentional non-determinism. That is, $puts$ and $gets$ can happen in any order, without changing the value that a program evaluates to.
\end{itemize}
%\begin{center}
%\textit{Monotonically increasing writes + threshold reads = deterministic parallelism}
%\end{center}
The resulting model is general enough to subsume the existing single-assignment model. The user-specified set just needs to be a \textit{join-semilattice}, which is a partially ordered set in which every two elements have a least upper bound.

@@

$Par$ type constructor($\ast$ is the normal haskell type):

@@

Internally, the $Par$ monad represents computations in continuation-passing style, in terms of their interpretation in the IO monad.

@@

In an attempt towards creating better deterministic parallel programming models, Lvish has extended the existing theme of monotonicity to provide better applicability, usability, performance and a single generic model that subsumes many of the single assignment models, thereby representing a larger class of problems which can be efficiently parallelized. The performance with regards to certain benchmarks are an evidence of the same. At the end of the previous section, some of the drawbacks of LVish and a proposed enhancement were highlighted from a recent paper.
